---
title: "Polling Simulation: Sampling and the Central Limit Theorem"
---

In this activity, you'll use data from the 2024 Cooperative Election Study (CES) — a large-scale national survey conducted during the 2024 election — to simulate what happens when a news organization tries to estimate public opinion from a poll sample. The full dataset has nearly 55,000 respondents, which we'll treat as our population.

Your goal is to see firsthand how sample size affects the accuracy of estimates, and what the distribution of sample estimates actually looks like — the core idea behind the Central Limit Theorem.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from this URL and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

We'll focus on one question: what share of American adults identify as religiously unaffiliated (atheist, agnostic, or nothing in particular)? This has been a major story in American religion journalism for years.

First, we'll create a new column called `unaffiliated` that is `TRUE` if a respondent's `trad` value is `"Atheist"`, `"Agnostic"`, or `"Nothing in Particular"`, and `FALSE` otherwise. Then calculate the proportion of the full dataset that is unaffiliated. Save this value as `population_proportion`.

```{r}
ces <- ces |>
  mutate(unaffiliated = trad %in% c("Atheist", "Agnostic", "Nothing in Particular"))

proportion <- mean(ces$unaffiliated)
```

That population proportion is the "true" population value that any poll is trying to estimate. Write it down — you'll be comparing everything else to it.

### Task 4: Take a single small sample

Use `set.seed(42)` and then sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of unaffiliated respondents in your sample.

```{r}
set.seed(42)
sample_50 <- ces |> sample_n(50)
mean(sample_50$unaffiliated)
```

How close is your sample estimate to the original proportion? Is it surprising?


ANSWER HERE: The original proportion is about .36, and my sample estimate is .32, so they are only about 4 percentage points apart. I do not really find this surprising because the sample only has 50 people, so I think some differences from the full population should be expected because of random chances. I would be more skeptical of seeing the 32% as a major change because of how small the sample is.


### Task 5: Take a single larger sample

Now sample 500 rows. Set a different seed value, grab the sample and save it as `sample_500`. Calculate the proportion of unaffiliated respondents in this sample.

```{r}
set.seed(405)
sample_500 <- ces |> sample_n(500)
mean(sample_500$unaffiliated)
```

How did the estimate change compared to the sample of 50? What does this suggest about sample size?


ANSWER HERE: The estimate has increased to .36 with the sample of 500 compared to the previous .32 with the sample of 50. The larger sample is a lot closer to the true population proportion which is about 36% This to me suggests that larger samples tend to give a more accurate estimate because they are not as affected by a random chance. 


### Task 6: Simulate many polls

A single sample can get lucky or unlucky. To understand how polling actually works, we need to see the *distribution* of estimates across many samples. The code below draws 1,000 samples of size 50, calculates the proportion of unaffiliated respondents in each, and stores the results.

```{r}
set.seed(150)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$unaffiliated)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side — one for the 1,000 samples of size 50 and one for the 1,000 samples of size 500. Add a vertical line at `proportion` to each so you can see how well the samples are centering on the true value.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously Unaffiliated",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**What shape do both histograms have?** Describe them in plain language. Does it surprise you that this shape emerges from a variable that is simply TRUE or FALSE for each respondent?


ANSWER HERE: Both histograms have a sort of bell-shaped distribution. Most of the sample estimates look like they are grouped around 36%, while fewer estimates are very high or low. The sample of 500 looks like it is much more tighter than the sample of 50. It is a little surprising that this shape comes from a variable that is only True or False, but when we take many random samples, I think the sample tends to form this bell-shaped pattern. 


**Compare the two histograms.** What is the most visible difference between the n=50 and n=500 distributions? What does that difference tell you about what happens when a poll uses a larger sample?


ANSWER HERE: The biggest difference that I notcied is that the n=500 histogram is a lot more narrower and tightly grouped around the true proportion, and the n=50 histogram is much more spread out. This to me shows that larger samples offer a more consistent or reliable estimate and are less affected by something like a random chance. 


**Both distributions are centered near the red line.** What does it mean that 1,000 different random samples — each one drawn independently — tend to cluster around the same value? What would it mean for journalism if that *weren't* the case?


ANSWER HERE: Because the 1,000 different random samples cluster around the same values to me shows that random sampling generally gives a reliable estimate of the true population. I think individual polls may be a little higher or lower. However, overall, I would say they center around the true value. Iff this was not the case I think it would be hard for journalists to trust polls or use samples to report what a larger group would think.


**A reporter tells you their story is based on a poll of 50 people.** Based on what you've seen in the simulation, what questions would you want to ask before trusting that number? You don't need to use any technical terms — just describe what you observed.


ANSWER HERE: The one thing I would want to know is how the 50 people were chosen. I also would want to know if they represent the larger population. I would want to ask how much the results might change if they surveyed a different group of 50 people. From the  simulation, I found that the samples of 50 had a wider range of results, so I would be careful about treating one small poll as the exact answer.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
