---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv('https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv') 
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: The data shows that there are 411 white house staff members. It also includes information about their name, employment, salary, pay and position. Salaries have a wide range depending on the position. Many of the higher positions are earning around $175,000 to $197,200, while lower-level positions such as staff assistants and coordinators tend to earn much less. From a reporting standpoint, what stands out to me is the large range in salaries and how pay appears to be different for each job based on title and responsibility. One issue that I think would create problems is that some salaries are listed at $0. I noticed that several advisors and senior officials have a zero-dollar salary. This is interesting because they have a high-level position but are not getting paid for it. These values could also lower the average salary and make it wrong and misleading, so I want to know why they are listed this way.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY), median_salary = median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY),median_salary = median(SALARY),min_salary = min(SALARY),max_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE: From my coding, I was able to find the mean salary, which is $112,703, and the median salary is $109,243. This shows me that the average salary is a little bit higher than the middle salary. The salaries range from zero to the highest number, which is 197,200. The zero-dollar salaries stand out to me a lot and could affect the mean. Because of this, I would want to look more into why some employees are listed with no salary before looking at anything else.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE: If I would have to pursue a story right now, I would be interested in looking more into the significant pay differences among white house staff members. The salaries ranging from $0 to $197,200 immediately caught my attention. I would like to look into why people who have a higher-ranked position are listed with a $0 salary while others are earning close to $200,000. I think my next step would be to look more into the employees who have the $0 salaries and see why they are unpaid in their positions, and compare them to other positions and employment status to the highest-paid workers.
