---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |> summarize(mean = mean(danceability), sd = sd(danceability)) 
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |> ggplot() + geom_histogram(aes(x = danceability), binwidth = .08)
```

### Task 5: Add a line to the histogram for the mean danceability score

Write code to add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |> ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.08) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1)
```

### Task 6: Describe our results

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and distribution tell us about the songs in the dataset?


After looking at the histogram, what is drawing my attention the most is that a majority of the songs have danceability scores between about .5 and .9, and the largest group is around .7 to .8. I think the distribution is slightly skewed to the left because there are a smaller number of songs with a lower danceability score that stretches over the left side of the histogram. I do not think it looks perfectly normal because the two sides are not perfectly even. The mean to me looks like it is around.67 which is represented by the dashed red line. This tells me that the average song in the data has a pretty high danceability score. I think it is interesting that there are few songs with very low danceability scores, but a lot of the songs are grouped toward the higher end. This observation is making me wonder if popular songs on Spotify are more danceable in general or if there is something about the songs that are in this dataset that causes the score to be higher.


### Task 7: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
