---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```


### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`


```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.


```{r}
ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- mean(ces$evangelical)

```


Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?


ANSWER HERE: The White Evangelical proportion is about 0.105, or 10.5%. This is much lower than the unaffiliated proportion that is 0.36 or 36%. To me, this means that in this dataset, unaffiliated respondents make up more than three times the share of the White Evangelical respondents. 


### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.


```{r}
set.seed(42)
sample_50 <- ces |> sample_n(50)
mean(sample_50$evangelical)
```


How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?


ANSWER HERE: My sample estimate was 0.16 or 16%, compared with the population proportion of about 10.5%, so it is around 5.5 percent higher. This is more of a bigger difference than what I noticed with the unaffiliated group, where the sample was 32% compared with about 36%. This to me shows that a sample of only 50 people can vary from the true population just because of random chance.


### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.


```{r}
set.seed(405)
sample_500 <- ces |> sample_n(500)
mean(sample_500$evangelical)
```


Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?


ANSWER HERE: I do think that the larger sample improved the estimate. With 50 people, the estimate was 16%, but with 500 people, it was 12%. This is a lot closer to the true population proportion of about 10.5%. I think this is similar to what we saw in class with the unaffiliated group. The increasing in the sample size also brought the estimate much closer to the true population value. I think this shows that larger samples generally give more accurate estimates.


### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.


```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```


### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.


```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share White Evangelical",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```


### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?


ANSWER HERE: I noticed that the n = 50 histogram is somewhat bell-shaped, but it is less symmetric and more spread out than the unaffiliated histogram from class today. Since White Evangelicals make up only about 10% of the population, I think a sample of 50 may have only a few White Evangelical respondents. For me, this means that small changes in who gets randomly selected can have a bigger effect on the sample proportion, I think this would cause more variation in the results.


**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?


ANSWER HERE: What I noticed is that the n = 500 histogram is much more narrower and more tightly grouped around the true proportion than the n = 50 histogram is. This is showing that increasing the sample size has a big impact on reducing the amount of variation between samples. I think the improvement appears bigger for the White Evangelical group because it makes up a smaller share of the population. Because of this, I think small samples can vary more.


**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?


ANSWER HERE: Because White Evangelicals make up about 10.5% of the population, I think a poll of 200 people would likely include about 21 White Evangelical respondents. I think that this is a pretty small number of people to represent the entire subgroup. Also, the estimate could change a lot depending on who happens to be selected. This to me suggests that estimates for smaller subgroups are less reliable unless the overall poll has a large enough sample.


**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?


ANSWER HERE: Before I even report the 78% number, I would want to know how many of the 400 people were actually White Evangelical voters, because I know from the data that they are a smaller subgroup of the population. Also, I would want to know how the people were selected. I would need to know if the sample is actually representing the population. I also want to know if the result could change with a different sample. I think smaller subgroups can have more variation so I would need to be  careful about reporting 78% without knowing how reliable the estimate is.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
